Recursive partitioning of linear growth curve models

### 2023-01-29 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
further improvements for initialization and partitioning subsections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3250 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: ccc6feeffe593e508b23ae959a2680667248aac0

Previous publications on recursive partitioning of models with mixed effects find that initializing the random-effect estimates with zero yields accurate estimates of the subgroup memberships and thus of the final models  \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}. \cite{SelaySimo12} assessed the impact of different initialization values and found only minor differences that became decreased further with increasing sample size. In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data. With longitudinal data, however, random effects tend to be more pronounced: Repeated measures on the same subjects tend to be correlated more closely than observations nested within the same unit in cross-sectional data. If the random effects are sizable, the inital assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations. Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery and predictive accuracy. More specifically, that means the algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.

@@

The subgroup structure in Step~1 is estimated by a GLM tree using the general model-based recursive partitioning (MOB) algorithm by \cite{ZeilyHoth08}. Here, we give a short general overview and then comment on the aspects of the algorithm that may be particularly relevant for LGCMs. In the case of GLMs (with random effects held constant in an offset), the MOB algorithm cycles iteratively through the following steps:

\begin{enumerate}
\setlength\itemsep{0em}
\renewcommand{\labelenumi}{\alph{enumi})}
	\item Fit the GLM to all observations in the current subgroup. 
	\item Test for instability of the GLM parameters with respect to each of the partitioning variables.
	\item If there is some overall parameter instability, split the subgroup with respect to the partitioning variable associated with the highest instability.
	\item Repeat Steps (a) through (c) in each of the resulting subgroups.
\end{enumerate}

Parameter stability in Step~(b) is tested using the the \textit{scores} (gradient contributions) from the GLM fitted in Step~(a). Under correct specification of the model and mild regularity conditions, the scores have an expected value of 0. The parameter stability tests evaluate whether the scores fluctuate randomly around their mean of 0, or exhibit systematic deviations when ordered by the values of a covariate available for partitioning. For continuous covariates $U_k$ (or ordered covariates with a large enough number of unique values), this involves computing the following cumulative score process $W_k(t)$ with respect to each potential partitioning variable \citep{ZeilyHoth08}:
%
\begin{eqnarray}
\label{eq:efp}
W_{k}(t) = \hat{J}^{-1/2} n_{j}^{-1/2} \sum^{[n_jt]}_{i=1}{\hat{\psi}}_{\sigma(U_{ik})}
\end{eqnarray}
%
where $\hat{J}$ is a suitable estimate of the covariance matrix of the parameter estimates, and $n_j$ gives the number of observations in the current subgroup. Further, $\hat{\psi}_{\sigma(U_{ik})}$ denotes the scores evaluated at the parameter estimates, with subscript {$\sigma(U_{ik})$} denoting their ordering by the values of partitioning variable $U_k$. Note that $0 \leq t \leq 1$, thus $n_jt = 1$ for an observation associated with a unique minimum on the partitioning variable, and $n_jt = n_j$ for an observation with a unique maximum.

>>>>

Previous publications on recursive partitioning of models with mixed effects find that initializing the random-effect estimates with zero yields accurate estimates of the subgroup memberships and thus of the final models  \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}. \cite{SelaySimo12} assessed the impact of different initialization values and found only minor differences that decreased further with increasing sample size. In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data. With longitudinal data, however, random effects tend to be more pronounced: Repeated measures on the same subjects tend to be correlated more closely than observations nested within the same unit in cross-sectional data. If the random effects are sizable, the inital assumption of zero random effects could provide an unrealistic starting point that may be difficult to overcome in subsequent iterations. Our expectation is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure may improve subgroup recovery and predictive accuracy. More specifically, that means the algorithm starts by estimating the classic version of the mixed-effects model from Equation~\ref{eq:GLMMtree} with just one set of fixed-effects coefficients $\beta$ and all subjects in a single group.
The alternative initialization step is thus:
%
\begin{enumerate}
\setlength\itemsep{0em}
\renewcommand{\labelenumi}{\arabic{enumi}$'$.}
	\setcounter{enumi}{-1}
	\item Initialize by setting step $r = 0$ and fit the mixed-effects model $g(\mu_{i}) = X_{i} \beta + Z_{i} b_{i, (r)}$ to the full sample. Extract the random-effect estimates~$\hat{b}_{i,(r)}$ from the fitted model.
\end{enumerate}

@@

The subgroup structure in Step~1 is estimated by a GLM tree using the general model-based recursive partitioning (MOB) algorithm of \cite{ZeilyHoth08}. Here, we give a short general overview and then comment on the aspects of the algorithm that may be particularly relevant for LGCMs. In the case of GLMs (with random effects held constant in an offset), the MOB algorithm cycles iteratively through the following steps: 
%
\begin{enumerate}
\setlength\itemsep{0em}
\renewcommand{\labelenumi}{(\alph{enumi})}
	\item Fit the GLM to all observations in the current subgroup. 
	\item Test for instability of the GLM parameters with respect to each of the partitioning variables.
	\item If there is some overall parameter instability, split the subgroup with respect to the partitioning variable associated with the highest instability.
	\item Repeat Steps (a) through (c) in each of the resulting subgroups.
\end{enumerate}
%
Parameter stability in Step~(b) is tested using the the \textit{scores} (gradient contributions) from the GLM fitted in Step~(a). Under correct specification of the model and mild regularity conditions, the scores have an expected value of 0. The parameter stability tests evaluate whether the scores fluctuate randomly around this mean of 0, or exhibit systematic deviations when ordered by the values of a covariate available for partitioning. For continuous covariates $U_k$ (or ordered covariates with a large enough number of unique values), this involves computing the following cumulative score process $W_k(t)$ with respect to each potential partitioning variable \citep{ZeilyHoth08}:
%
\begin{eqnarray}
\label{eq:efp}
W_{k}(t) = \hat{J}^{-1/2} n_{j}^{-1/2} \sum^{[n_jt]}_{i=1}{\hat{\psi}}_{\sigma(U_{ik})}
\end{eqnarray}
%
where $\hat{J}$ is a suitable estimate of the covariance matrix of the parameter estimates, and $n_j$ gives the number of observations in the current subgroup. Further, $\hat{\psi}_{\sigma(U_{ik})}$ denotes the scores evaluated at the parameter estimates, with subscript {$\sigma(U_{ik})$} denoting their ordering by the values of partitioning variable $U_k$. Note that $0 \leq t \leq 1$, thus $n_jt = 1$ for an observation associated with a unique minimum on the partitioning variable, and $n_jt = n_j$ for an observation with a unique maximum.
